---
title: "hm02-gapminder exploration"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this Rmd file, I will be exploring the `gapminder` dataset using `ggplot2`. To get started, I will be loading the `tidyverse` package and the `gapminder` dataset.


```{r preliminaries}
library(tidyverse)
library(gapminder)
```


## Smell test the data

### The `gapminder` dataset

The `gapminder` dataset contains data on life expectancy, GDP per capita and population by country. We can see the head of the dataset:

```{r head}
head(gapminder)
```

We can check that `gapminder` is dataframe:

```{r inherits}
inherits(gapminder,'data.frame')
```

As we can see below, `gapminder` is a dataframe which has has 1704 rows and 6 columns. 

```{r basic dataframe info}
typeof(gapminder)
nrow(gapminder)
ncol(gapminder)
```

We can also use the `dim` function to get the dimensions of the dataframe:

```{r dimensions}
dim(gapminder)
```

From the below, we can see that two of the columns are categorical variables and the remaining four are continuous random variables.

```{r summary}
summary(gapminder)
```

We can use the `sapply` function to check the type and class of each of the columns of the `gapminder` dataframe:

```{r column types}
sapply(gapminder,typeof)
sapply(gapminder,class)
```

In particular, notice that the summary shows that `year` is being treated as a continuous variable. We can make `R` to treat it as a categorical variable thus:
```{r as.factor}
gapminder$year = as.factor(gapminder$year)
class(gapminder$year)
summary(gapminder$year)
```

## Explore individual variables

### Categorical variable

We can have a look at the countries categorical variable. We can select the countries column from `gapminder` using `select`:

```{r country}
select(gapminder,country)
```

This has a lot of replicates due to multiple years' worth of data. We can remove the duplicates using the `unique` function in `R`:

```{r unique countries}
unique(select(gapminder,country))
```

Then we can find the total number of countries in the data set using the `count` function:

```{r countries count}
gapminder %>% 
  select(country) %>% 
  unique() %>% 
  count()
```

We can find the number of times each country has appeared on this dataset as follows (only first 6 shown):

```{r countries reps}
country_count = table(gapminder$country)
head(country_count)
```

We can also look at how many countries (in this dataset) are in each continent:

```{r countries bar plot}
gapminder %>% 
  select(country,continent) %>% 
  unique() %>% 
  ggplot(aes(x=continent)) +
  geom_bar(aes(fill=continent)) +
  ggtitle('Countries Per Continent')
```

### Quantitative variable

Let's take a look at the GDP per capita data. Here's a quick summary of the data:

```{r gdp summary table}
gapminder %>% 
  select(gdpPercap) %>% 
  summary()
```

We can see this on a basic box plot. From the summary, we can see the average and median are very log compared to the maximum, so we will use a log scale in the box plot:

```{r gdp boxplot}
ggplot(gapminder,aes(y=gdpPercap)) + 
  scale_y_log10() +
  geom_boxplot()
```

We can also look at the GDP based on continent:

```{r gdp boxplot continent}
ggplot(gapminder,aes(x=continent,y=gdpPercap)) + 
  scale_y_log10() +
  geom_boxplot()
```

We can also see the spread of the data from a histogram. Similar to before, we used a log scale, but this time for the x axis.

```{r gdp histogram}
ggplot(gapminder,aes(gdpPercap)) +
  geom_histogram(bins=50) +
  scale_x_log10()
```

We can combine the histogram and a density plot:

```{r gdp histogram density}
ggplot(gapminder,aes(gdpPercap)) +
  scale_x_log10() +
  geom_histogram(aes(y=..density..),alpha=0.5) +
  geom_density()
```

## Explore various plot types

Let's compare GDP per capita against population on a scatter plot:

```{r pop vs gdp}
ggplot(gapminder,aes(pop,gdpPercap)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_point(alpha=0.3)
```

We can see the same plot but with different continents reflected as different colours:

```{r pop vs gdp colour}
ggplot(gapminder,aes(pop,gdpPercap)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_point(aes(colour=continent),alpha=0.4)
```

We can also use colours to reflect which points has a high life expectancy:

```{r pop vs gdp with life exp}
ggplot(gapminder,aes(pop,gdpPercap)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_point(aes(colour=lifeExp>65),alpha=0.4)
```

Lastly, we can track how a certain country's GDP and population changed on the above plot. For this illustation, I decided to choose Japan I used the `last_plot()` command to pull up the previous plot.

```{r pop vs gdp with path}
last_plot() + 
  geom_path(data=filter(gapminder,country=='Japan'),arrow=arrow())
```

